%% Elsevier CAS single-column manuscript for double-anonymized review
%% Compile with: pdflatex main_els_anonymous; bibtex main_els_anonymous; pdflatex main_els_anonymous; pdflatex main_els_anonymous

\documentclass[a4paper,fleqn]{cas-sc}

%% APA-style author-year citations (Information Processing & Management uses APA).
%% If the CAS class rejects natbib author-year for some reason, fall back to
%% \usepackage[numbers,sort&compress]{natbib} and \bibliographystyle{cas-model2-names}.
\usepackage[authoryear,round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{float}
\usepackage{afterpage}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,calc}
\usepackage{tabularx}
\usepackage{array}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{url}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage{ragged2e}

%% Typography: cas-sc.cls already loads \usepackage[T1]{fontenc} (line 144) and
%% the STIX/Charis text fonts, so we do NOT re-load fontenc, and we deliberately
%% skip lmodern (it would override the class-provided STIX/Charis faces).
\usepackage{microtype}        % font expansion + kerning to fix bad line breaks
\emergencystretch=2em         % last-resort stretch before overfull hboxes

\newcolumntype{Y}{>{\RaggedRight\arraybackslash}X}
\newcommand{\na}{\textemdash}

\hypersetup{
  pdfauthor={},
  pdftitle={Speech Large Language Models in the Real World: A Survey of Training, Evaluation, Deployment, and Responsible Design},
  pdfsubject={},
  pdfkeywords={}
}

\newcommand{\cmark}{\checkmark}
\newcommand{\xmark}{\ding{55}}

\begin{document}
\let\WriteBookmarks\relax
\def\floatpagepagefraction{1}
\def\textpagefraction{.001}

\shorttitle{Speech Large Language Models in the Real World}
\shortauthors{Anonymous Author(s)}

\title[mode=title]{Speech Large Language Models in the Real World: A Survey of Training, Evaluation, Deployment, and Responsible Design}

%% Double-anonymized review version: author identities, affiliations,
%% emails, ORCID IDs, acknowledgements, and self-identifying URLs are omitted.
% \author[1]{Anonymous Author(s)}
% \affiliation[1]{organization={Anonymous Institution},
%                 city={Anonymous City},
%                 country={Anonymous Country}}
\author{Anonymous Author(s)}
\begin{abstract}
Speech Large Language Models (Speech LLMs) are moving into real-time
voice assistants and full-duplex systems, yet existing surveys mainly
explain how these models are built rather than whether they can be
evaluated, deployed, and governed in real-world settings. We audit the
field through a lifecycle lens spanning training, evaluation,
deployment, and responsible design. Our corpus includes 18 Speech LLM
benchmarks released between 2024 and June 2026 (3 peer-reviewed, 15
arXiv preprints), coded across 17 capability dimensions and four
validity perspectives into an 18$\times$17 coverage matrix. The audit
identifies four gaps. First, benchmark coverage remains skewed toward
ASR, spoken question answering, and emotion (12/18 each), while safety
(4/18), tool use (2/18), latency (1/18), and privacy (0/18) are rarely
evaluated. Second, no benchmark occupies the
Use-case$\times$Capability cell in our scope--lens taxonomy. Third,
six interaction-first systems (Moshi, GPT-4o-Audio, MiniCPM-o~4.5,
Voila, OmniFlatten, Gemini~Live) appear in at most one public
benchmark. Fourth, five documented cross-stage couplings show that
training, evaluation, deployment, and safety are interdependent. We
synthesize these findings into benchmark design principles,
Speech-LLM Evaluation and Deployment Card templates, and a
speech-specific governance framework for audio jailbreaks, voice
cloning, speaker re-identification, and provenance under
transformation.
\end{abstract}

\begin{highlights}
\item We audit 18 Speech LLM benchmarks and show that privacy (0/18), latency (1/18), and safety (4/18) remain under-evaluated.
\item A scope-by-lens taxonomy reveals no benchmark that jointly evaluates realistic use cases and specific speech capabilities.
\item A reverse model audit shows that six interaction-first systems are absent from most public benchmark records.
\item Five cross-stage couplings show how training choices propagate into evaluation blind spots, deployment failures, and safety risks.
\item We propose evaluation and deployment reporting cards that connect accuracy, latency, streaming behavior, safety, and governance.
\end{highlights}

\begin{keywords}
Speech large language models \sep Audio language models \sep Multimodal evaluation \sep Streaming inference \sep Responsible AI \sep Survey
\end{keywords}

%% Suppress the empty "ORCID(s):" footnote line the class emits by default.
%% \printorcid is defined in cas-common.sty; we redefine (not modify) it here.
\let\printorcid\relax

\maketitle

\input{sections/SECTION1_INTRO}
\input{sections/SECTION2_PRELIMINARY}
\input{sections/SECTION3_TRAINING}
\input{sections/SECTION4_EVALUATION}
\input{sections/SECTION5_DEPLOYMENT}
\input{sections/SECTION6_SAFETY}
\input{sections/SECTION7_FUTURE}
\input{sections/SECTION9_DISCUSSION}
\input{sections/SECTION8_CONCLUSION}

%% IPM Guide for Authors: this declaration is required immediately before
%% the references list. Wording follows the template in the IPM guide.
\FloatBarrier
\section*{Declaration of generative AI and AI-assisted technologies in the manuscript preparation process}
During the preparation of this work, the authors used generative AI tools (large language models) to support language polishing, structural revision, and checklist-based editing of drafts. After using these tools the authors reviewed and edited the content as needed and take full responsibility for the content of the published article.

\section*{Data Availability}
The benchmark coding matrix and supplementary reading notes supporting this survey are provided as supplementary material.

\bibliographystyle{elsarticle-harv}
\bibliography{references}

\end{document}
